Windowed least-squares model reduction for dynamical systems

### 2019-10-05 main.tex
name change
reread intro

%\title{Model Reduction via \methodName}
%\title{Windowed least--squares reduced-order models for dynamical systems}
\title{Windowed residual-minimization model reduction for dynamical systems}

@@

\address[a]{Sandia National Laboratories,  Livermore, CA}
\begin{abstract}
This work proposes a \methodNameLower\ (\methodAcronym) model-reduction
	formulation for dynamical systems. The proposed approach sequentially minimizes the
	time-continuous full-order-model residual over a low-dimensional trial
	subspace across time windows. The approach comprises a generalization 
of existing model reduction approaches; we show that various instances of
  the method recover the Galerkin,
  least-squares Petrov--Galerkin (LSPG), and space–time LSPG approaches. The approach further addresses key deficiencies in
	existing model-reduction techniques, e.g., the dependence of (space--time) LSPG 
on the time-discretization and the exponentially growing \textit{a
	posteriori} error bounds present in Galerkin and
	LSPG.
We consider two types of trial subspaces
	within the proposed framework:
	one that reduces only the spatial dimension of the full-order model, and one that
	reduces both the spatial and temporal dimensions of the full-order model.
	For each type of trial subspaces, we consider two different solution
	techniques: direct (i.e., discretize-then-optimize) and indirect (i.e.,
	optimize-then-discretize). %Critically, we show that various instances of
	%this framework yield existing model-reduction approaches, namely Galerkin,
	%LSPG, and space–time LSPG projection. 
  Numerical experiments are presented for trial subspaces with spatial dimension reduction, in where it is demonstrated that the \methodAcronym\ formulation can yield more
	accurate and physically relevant solutions than Galerkin and
	LSPG projection.

@@

Simulating parameterized dynamical systems underpins many applications across
science and engineering, including engineering design, control, financial
analysis, and epidemiology. In many contexts, executing a dynamical-system
simulation at a single parameter instance---which entails the numerical
integration of a system of ordinary differential equations (ODEs)---incurs an
extremely large computational expense.  This occurs, for example, when the
state-space dimension is large (e.g., due to fine spatial resolution when
discretizing a partial differential equation) and/or when the number of time
instances is large (e.g., due to time-step limitations incurred by stability
or accuracy considerations).  When the application is time critical or many
query in nature, analysts must replace such large-scale parameterized
dynamical-system models (which we refer to as the full-order model) with a low-cost approximation that makes the
application tractable.

@@

Projection-based reduced-order models (ROMs) comprise one such approximation
strategy. First, these techniques execute a computationally expensive
\textit{offline} stage that computes a low-dimensional ``trial'' subspace on
which the dynamical-system state can be well approximated (e.g., by computing
state ``snapshots'' at different time and parameter instances, by solving
Lyapunov equations, etc). Second, these methods execute an inexpensive
\textit{online} stage during which they compute approximations to the
dynamical-system trajectory that reside on this trial subspace via a
projection process.

Model reduction for linear-time-invariant systems (and other strongly
structured dynamical systems) is quite mature \cite{wilcox_benner_rev,moore,roberts,GugercinIRKA} as system-theoretic properties (e.g., controllability,
observability, asymptotic stability, $\mathcal H_2$-optimality) can be readily
quantified and accounted for in the projection process, which yields
reduced-order models that inherit such properties.  The primary challenge for
developing reduced-order models for general nonlinear dynamical systems is
that such properties are difficult to quantitatively assess. As a result, it
is difficult to develop reduced-order models that preserve important
dynamical-system properties, which often results in inaccurate or unstable
methods.  To address this, researchers have pursued several directions that
aim to imbue reduced-order models for nonlinear
dynamical systems with properties that can improve robustness and accuracy.
These
efforts include residual-minimizing
projection that equips the ROM solution with a notion of optimality~\cite{carlberg_lspg,carlberg_gnat,legresley_1,legresley_2,legresley_3,bui_resmin_steady,bui_unsteady,rovas_thesis,carlberg_thesis,bui_thesis,l1};
space--time
projection that leads to error bounds that grow slowly in time~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,benner_st};
``energy-based'' inner
products that ensure non-increasing entropy in the ROM solution~\cite{rowley_pod_energyproj,Kalashnikova_sand2014};
basis-adaptation methods that improve the ROM's accuracy \textit{a
posteriori}~\cite{carlberg_hadaptation,adeim_peherstorfer,etter2019online}, stabilizing subspace rotations that account for truncated modes \textit{a priori}~\cite{basis_rotation}, structure-preserving
methods that enforce conservation \cite{carlberg_conservative_rom} or
(port-)Hamiltonian/Lagrangian structure
\cite{LALL2003304,carlberg2012spd,structurePreserveBeattie,chaturantabut2016structure,farhat2014dimensional} in the ROM; and subgrid-scale
modeling methods that aim to improve accuracy by addressing the closure
problem~\cite{san_iliescu_geostrophic,iliescu_pod_eddyviscosity,iliescu_vms_pod_ns,Bergmann_pod_vms,iliescu_ciazzo_residual_rom,Wang_ROM_thesis,wentland_apg,Wang:269133,San2018}.
We note that these techniques are often not mutually exclusive.
%Residual-minimizing projection methods are the most relevant to the current work and comprise the focus of the following review.
Residual-minimizing and space--time projection are the most relevant classes of
methods to the current work and comprise the focus of the following review.

@@

Residual minimization methods in model-order reduction comprise 
computing a solution within a low-dimensional trial subspace that minimizes some residual norm of the system of interest. Researchers have developed such residual-minimizing
model-reduction methods for both static systems (i.e., systems without
time-dependence)
\cite{legresley_1,legresley_2,legresley_3,bui_resmin_steady,rovas_thesis,carlberg_thesis,bui_thesis}
and dynamical
systems~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg,carlberg_lspg_v_galerkin}.
In the latter category,
Refs.~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg}
formulated the residual minimization problem for dynamical systems by
sequentially minimizing the \textit{time-discrete} full-order-model residual
(i.e., the residual arising after applying time discretization) at each time
instance on the time-discretization grid. This formulation is often referred
to as the \textit{least--squares Petrov--Galerkin} (LSPG) method.
Ref.~\cite{carlberg_lspg_v_galerkin} performed detailed analyses of this
formulation and examined its connections with Galerkin projection. Critically,
this work demonstrated that (1) under certain conditions, LSPG can be derived
from a Petrov--Galerkin projection executed on the time-continuous
full-order-model residual, and (2) LSPG and Galerkin projection are equivalent
in the limit as the time step goes to zero (i.e., Galerkin projection
minimizes the time-instantaneous full-order model residual). Numerous
numerical experiments have demonstrated that LSPG projection often yields more
accurate and stable solutions than Galerkin
projection~\cite{bui_thesis,carlberg_lspg_v_galerkin,carlberg_gnat,carlberg_thesis,parish_apg}.
The common intuitive explanation for this improved performance is that, by
minimizing the full-order-model residual over a finite time window (rather
than time instantaneously), LSPG projection computes solutions that are more
accurate over a larger part of the trajectory as compared to Galerkin projection.
%\KTC{Let's discuss. What specifically do we want to point out about
%deficiencies in LSPG in this section? Should we also look in ST-LSPG? It seems you're trying
%to say that we don't like how the time step size \textit{and} time
%discretization method dictate so much about the method (I agree with this). If
%so, we should structure this to include other discretize-then-optimize methods
%I think. Or we could just focus on the time-step aspect (i.e., not the time
%discretization method), in which case we should focus on LSPG.}
%\KTC{HERE}

@@

%\section*{References}
\bibliographystyle{aiaa}
\bibliography{refs}{}
\end{document}

>>>>

%\title{Model Reduction via \methodName}
%\title{Windowed least--squares reduced-order models for dynamical systems}
\title{Windowed least-squares model reduction for dynamical systems}

@@

\address[a]{Sandia National Laboratories,  Livermore, CA}
\begin{abstract}
This work proposes a \methodNameLower\ (\methodAcronym) model-reduction
	formulation for dynamical systems. The proposed approach sequentially minimizes the
	time-continuous full-order-model residual over a low-dimensional space--time trial
	subspace defined over an arbitrarily defined time window. The approach comprises a generalization 
of existing model reduction approaches, as particular instances of
  the methodology associate with Galerkin,
	least-squares Petrov--Galerkin (LSPG), and space–time LSPG projection. In
	addition, the approach
	addresses key deficiencies in existing model-reduction
	techniques, e.g., the dependence of (space--time) LSPG projection on the
	time discretization and the exponential growth in time exhibited by \textit{a posteriori}
	error bounds for Galerkin and LSPG projection.  We consider two types of
	space--time trial
	subspaces within the proposed framework: one that reduces only the spatial
	dimension of the full-order model, and one that reduces both the spatial and
	temporal dimensions of the full-order model.  For each type of trial
	subspace, we consider two different solution techniques: direct (i.e.,
	discretize-then-optimize) and indirect (i.e., optimize-then-discretize).
	Numerical experiments conducted using the trial subspaces characterized by spatial
	dimension reduction demonstrate that the \methodAcronym\
	approach can yield more accurate and physically relevant solutions \KTC{What
	does `physically relevant' mean? This is imprecise.} than
	Galerkin and LSPG projection.

@@

Simulating parameterized dynamical systems arises in many applications across
science and engineering, including design, control, uncertainty
quantification, and epidemiology, for example. In many contexts, executing a dynamical-system
simulation at a single parameter instance---which entails the numerical
integration of a system of ordinary differential equations (ODEs)---incurs an
extremely large computational cost.  This occurs, for example, when the
state-space dimension is large (e.g., due to fine spatial resolution when
discretizing a partial differential equation) and/or when the number of time
instances is large (e.g., due to time-step limitations incurred by stability
or accuracy considerations).  When the application is time critical or many
query in nature, analysts must replace such large-scale parameterized
dynamical-system models (which we refer to as the full-order model) with a low-cost approximation that makes the
application tractable.

@@

Projection-based reduced-order models (ROMs) comprise one such approximation
strategy. First, these techniques execute a computationally expensive
\textit{offline} stage that computes a low-dimensional \textit{trial subspace} on
which the dynamical-system state can be well approximated (e.g., by computing
state ``snapshots'' at different time and parameter instances, by solving
Lyapunov equations). Second, these methods execute an inexpensive
\textit{online} stage during which they compute approximations to the
dynamical-system trajectory that reside on this trial subspace via a
projection process.

Model reduction for linear-time-invariant systems (and other well structured
dynamical systems) is quite mature
\cite{wilcox_benner_rev,moore,roberts,GugercinIRKA}, as system-theoretic properties (e.g., controllability,
observability, asymptotic stability, $\mathcal H_2$-optimality) can be readily
quantified and accounted for in the projection process; this often results in
reduced-order models that inherit such important properties.  The primary
challenge in
developing reduced-order models for general nonlinear dynamical systems is
that such properties are difficult to assess quantitatively. As a result, it
is difficult to develop reduced-order models that preserve important
dynamical-system properties, which often results in methods that yield
trajectories that are
inaccurate, unstable, or violate physical properties.  To address this, researchers have pursued several directions that
aim to imbue reduced-order models for nonlinear
dynamical systems with properties that can improve robustness and accuracy.
These
efforts include residual-minimizing
projection that equips the ROM solution with a notion of optimality~\cite{carlberg_lspg,carlberg_gnat,legresley_1,legresley_2,legresley_3,bui_resmin_steady,bui_unsteady,rovas_thesis,carlberg_thesis,bui_thesis,l1};
space--time
projection that leads to error bounds that grow slowly in time~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,benner_st};
``energy-based'' inner
products that ensure non-increasing entropy in the ROM
solution~\cite{rowley_pod_energyproj,Kalashnikova_sand2014,chan2019entropy};
basis-adaptation methods that improve the ROM's accuracy \textit{a
posteriori}~\cite{carlberg_hadaptation,adeim_peherstorfer,etter2019online}, stabilizing subspace rotations that account for truncated modes \textit{a priori}~\cite{basis_rotation}, structure-preserving
methods that enforce conservation \cite{carlberg_conservative_rom} or
(port-)Hamiltonian/Lagrangian structure
\cite{LALL2003304,carlberg2012spd,structurePreserveBeattie,chaturantabut2016structure,farhat2014dimensional} in the ROM; and subgrid-scale
modeling methods that aim to improve accuracy by addressing the closure
problem~\cite{san_iliescu_geostrophic,iliescu_pod_eddyviscosity,iliescu_vms_pod_ns,Bergmann_pod_vms,iliescu_ciazzo_residual_rom,Wang_ROM_thesis,wentland_apg,Wang:269133,San2018}.
We note that these techniques are often not mutually exclusive.
%Residual-minimizing projection methods are the most relevant to the current work and comprise the focus of the following review.
Residual-minimizing and space--time projection are the most relevant classes of
methods for the current work and comprise the focus of the following review.

@@

\KTC{HERE, COFFEE BREAK}
Residual minimization methods in model reduction comprise 
computing a solution within a low-dimensional trial subspace that minimizes some residual norm of the system of interest. Researchers have developed such residual-minimizing
model-reduction methods for both static systems (i.e., systems without
time-dependence)
\cite{legresley_1,legresley_2,legresley_3,bui_resmin_steady,rovas_thesis,carlberg_thesis,bui_thesis}
and dynamical
systems~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg,carlberg_lspg_v_galerkin}.
In the latter category,
Refs.~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg}
formulated the residual minimization problem for dynamical systems by
sequentially minimizing the \textit{time-discrete} full-order-model residual
(i.e., the residual arising after applying time discretization) at each time
instance on the time-discretization grid. This formulation is often referred
to as the \textit{least--squares Petrov--Galerkin} (LSPG) method.
Ref.~\cite{carlberg_lspg_v_galerkin} performed detailed analyses of this
formulation and examined its connections with Galerkin projection. Critically,
this work demonstrated that (1) under certain conditions, LSPG can be derived
from a Petrov--Galerkin projection executed on the time-continuous
full-order-model residual, and (2) LSPG and Galerkin projection are equivalent
in the limit as the time step goes to zero (i.e., Galerkin projection
minimizes the time-instantaneous full-order model residual). Numerous
numerical experiments have demonstrated that LSPG projection often yields more
accurate and stable solutions than Galerkin
projection~\cite{bui_thesis,carlberg_lspg_v_galerkin,carlberg_gnat,carlberg_thesis,parish_apg}.
The common intuitive explanation for this improved performance is that, by
minimizing the full-order-model residual over a finite time window (rather
than time instantaneously), LSPG projection computes solutions that are more
accurate over a larger part of the trajectory as compared to Galerkin projection.
%\KTC{Let's discuss. What specifically do we want to point out about
%deficiencies in LSPG in this section? Should we also look in ST-LSPG? It seems you're trying
%to say that we don't like how the time step size \textit{and} time
%discretization method dictate so much about the method (I agree with this). If
%so, we should structure this to include other discretize-then-optimize methods
%I think. Or we could just focus on the time-step aspect (i.e., not the time
%discretization method), in which case we should focus on LSPG.}
%\KTC{HERE}

@@

%\section*{References}
\bibliographystyle{siam}
\bibliography{refs}
\end{document}
